---
title: "HW5 - Make Data Move"
author: "Emilia Hansen"
date: '2020-10-19'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/emiliahansen/hw-instructions")
```

#  Explore the recent global developments with R
Today, you will load a filtered gapminder dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks below.**

## Get the necessary packages
First, start with installing the relevant packages 'tidyverse', 'gganimate', and 'gapminder'.

```{r libraries, echo = FALSE}
library(tidyverse)
library(gganimate)
library(gapminder)
```

Since R gives me the message that some other packages, should be installed, I will do so. 
```{r}
library(av)
library(gifski)
```

## Look at the data
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
unique(gapminder$year)
head(gapminder)
```
The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```
We see an interesting spread with an outlier to the right. Answer the following questions, please:

Q1. Why does it make sense to have a log10 scale on x axis?
answer: the scale_x_log10 function makes it possible to override the default linear mapping. Without the  log10 scale, the data in the plot would look cluttered and it would be difficult to read. An example of that is here below. 

```{r}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() 
```

Q2. What country is the richest in 1952 (far right on x axis)? 
answer: Kuwait
```{r}
gapminder_1952 <- gapminder %>%  #create a new object
  select(country, year, gdpPercap) %>% #select the needed columns to answer the question
  filter(year == "1952") %>% #filter by the year 1952
  arrange(desc(gdpPercap)) #arrange in descending order the data on gdp per capita

gapminder_1952[1, ] #use the object with brackets to get the first row in the arranged dataset
```
You can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```
The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

Q3. Can you differentiate the continents by color and fix the axis labels?
```{r}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) + #reusing the code from above to make the plot
  geom_point(aes(color = continent)) + #command the color to be shown on continent
  labs(x = "GDP per Capita") + #change the name of the x-axis
  labs(y = "Life Expectancy") + #change the name of the y-axis
  scale_x_log10() 
```

Q4. What are the five richest countries in the world in 2007?
Answer: Norway, Kuwait, Singapore, United States, 
```{r}
gapminder %>% # first I create a new object 
  select(country, year, gdpPercap) %>%  # Then I select the columns I need
  filter(year == "2007") %>% # I filter by the year 2007 
  arrange(desc(gdpPercap)) %>% # to sort the highest gdp per capita, I choose the arrange function with descent
  head(5) #I choose head to get the first rows and choose five to limit the shown countries to 5

```



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. And there are two ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim
```

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the 'Viewer' pane, not in this rmd preview. You need to knit the document to get the viz inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smoothes the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.


Q5 Can you add a title to one or both of the animations above that will change 
 in sync with the animation? [hint: search labeling for transition_states() and transition_time() functions respectively]
 
Making a title for anim1: transitions_states
```{r anim1 custom}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1) + 
  labs(title = "Year: {closest_state}") # make the title year and add in the function to show the closest state for the animation. 

```

Making a tite for  anim2: transitions_year
```{r anim2 custom}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year) +
  labs(title = "Year: {frame_time}") #Make the title "Year" and add the variable frame_time to show the years as done in the transistion_states animation
anim2
```

Q6 Can you make the axes' labels and units more readable? Consider expanding the abreviated labels as well as the scientific notation in the legend and x axis to whole numbers.[hint:search disabling scientific notation]
```{r anim2 custum}
options(scipen = 999) #disable the scientific notiation to make the units on the x-axis more readable. To prevent the scientific notation, a a large positive number needs to be used like 999
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year) +
  labs(title = "Year: {frame_time}") + #Make the title "Year" and add the variable frame_time to show the years as done in the transistion_states animation 
  labs(y = "Life Expectancy") + #Change the label for the y-axis to "Life Expectancy"
  labs(x = "GDP Per Capita") #Change the label for the x-axis to "GDP per Capita".
anim2
```


Q7 Come up with a question you want to answer using the gapminder data and write it down. Then, create a data visualisation that answers the question and explain how your visualization answers the question. (Example: you wish to see what was mean life expectancy across the continents in the year you were born versus your parents' birth years). [hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset and download more at https://www.gapminder.org/data/ ]

My question: what has Denmark's GDP per capita changed to over time? 

```{r}
gapminder_DK <- gapminder %>%  # I create the object gapminder_DK to ease the proces later on
  select(country, year, gdpPercap) %>% #I need the colums "country", "year" to observe the change over time, and "gdpPercap" to observe the changes
  filter(country == "Denmark") %>%  #I filter by Denmark, because that is the only country, I want to investigate.
  arrange(desc(gdpPercap)) #I arrange it by GDP per capita, and make them appaer from highest till lowest. 
gapminder_DK #showing the tibble of the development


```
Answer: going through the tibble, and can see that the development of GDP has been increasing each year. However, it is difficult to see if the development has been linear or whether it has been increasing for each each year, because there's a lot of data. To ease the interpretation, I will make a visualisation: 

```{r DK}
ggplot(subset(gapminder, country == "Denmark"), aes(year, gdpPercap)) + # Choose Denmark as the country to subset in the plot. Then choose year on the x-axis and gdpPercap on the y-axis 
  geom_point() 
```
The visualization shows that it has almost been an overall linear development of GDP per capita through time with a few dips in 1982 and 1992

